\documentstyle[11pt,reduce]{article}
\title{The Standard Lisp Report}
\date{}
\author{Jed Marti \\ A. C. Hearn \\ M. L. Griss \\ C. Griss}

%%% Function/method definition.
%%% de{fname}{arglist}{type}{text}          For short arg lists.
%%% DE{fname}{arglist}{type}{text}          For long arg lists.
\newlength{\argwidth}                  % Width of argument box.
\setlength{\argwidth}{4in}
\newlength{\dewidth}
\setlength{\dewidth}{4.5in}             % Width of text box.

\newcommand{\de}[4]
{\vspace{.25in} \noindent
\begin{minipage}[t]{\textwidth} \index{#1} {\f{#1}}{#2}\hfill{\em #3} \\
\hspace*{.25in}\begin{minipage}[t]{\dewidth} #4 \end{minipage}
\end{minipage} }

%%% Global/fluid variable description.
%%% variable{name}{initial value}{type}{text}
\newcommand{\variable}[4]
{\vspace{.25in} \noindent
\begin{minipage}[t]{\textwidth} \index{#1 (#3)} {\bf #1} = #2 \hfill {\em #3}
 \\
\hspace*{.25in} \ \begin{minipage}[t]{\dewidth} #4 \end{minipage}
\end{minipage}}

%%% Command to display an error or warning message in teletype format. Also
%%% leaves blank vertical space around it.
\newcommand{\errormessage}[1]
{\vspace{.1in} \noindent {\tt #1} \\ \vspace{.1in}}


%%% \p is a parameter name (or argument). Just do this as bf.
\newcommand{\p}[1] {{\bf #1}}
%%% \ty is a type - do as italics.
\newcommand{\ty}[1] {{\em #1}}
\begin{document}
\maketitle

\section{Introduction}
Although the programming language LISP was first formulated in
1960~\cite{LISP1.5}, a widely accepted standard has never appeared. As
a result, various dialects of LISP were
produced~\cite{CDC-LISP,LISP/360,MACLISP,Interlisp,LISPF1,LISP1.6} in
some cases several on the same machine! Consequently, a user often
faces considerable difficulty in moving programs from one system to
another. In addition, it is difficult to write and use programs which
depend on the structure of the source code such as translators,
editors and cross-reference programs.

In 1969, a model for such a standard was produced~\cite{Hearn:69} as
part of a general effort to make a large LISP based algebraic
manipulation program, REDUCE~\cite{REDUCE3.3}, as portable as
possible.  The goal of this work was to define a uniform subset of
LISP 1.5 and its variants so that programs written in this subset
could run on any reasonable LISP system.

In the intervening years, two deficiencies in the approach taken in
Ref.~\cite{Hearn:69} have emerged. First in order to be as general as
possible, the specific semantics and values of several key functions
were left undefined. Consequently, programs built on this subset could
not make any assumptions about the form of the values of such
functions. The second deficiency related to the proposed method of
implementation of this language. The model considered in effect two
versions of LISP on any given machine, namely Standard LISP and the
LISP of the host machine (which we shall refer to as Target LISP).
This meant that if any definition was stored in interpretive form, it
would vary from implementation to implementation, and consequently one
could not write programs in Standard LISP which needed to assume any
knowledge about the structure of such forms. This deficiency became
apparent during recent work on the development of a portable compiler
for LISP~\cite{PLC}. Clearly a compiler has to know precisely the
structure of its source code; we concluded that the appropriate source
was Standard LISP and not Target LISP.

With these thoughts in mind we decided to attempt again a definition
of Standard LISP. However, our approach this time is more aggressive.
In this document we define a standard for a reasonably large subset of
LISP with as precise as possible a statement about the semantics of
each function. Secondly, we now require that the target machine
interpreter be modified or written to support this standard, rather
than mapping Standard LISP onto Target LISP as previously.

We have spent countless hours in discussion over many of the
definitions given in this report. We have also drawn on the help and
advice of a lot of friends whose names are given in the
Acknowledgements. Wherever possible, we have used the definition of a
function as given in the LISP 1.5 Programmer's Manual~\cite{LISP1.5}
and have only deviated where we felt it desirable in the light of LISP
programming experience since that time. In particular, we have given
considerable thought to the question of variable bindings and the
definition of the evaluator functions EVAL and APPLY. We have also
abandoned the previous definition of LISP arrays in favor of the more
accepted idea of a vector which most modern LISP systems support.
These are the places where we have strayed furthest from the
conventional definitions, but we feel that the consistency which
results from our approach is worth the redefinition.

We have avoided entirely in this report problems which arise from
environment passing, such as those represented by the FUNARG problem.
We do not necessarily exclude these considerations from our standard,
but in this report have decided to avoid the controversy which they
create. The semantic differences between compiled and interpreted
functions is the topic of another paper~\cite{PLC}. Only functions
which affect the compiler in a general way make reference to it.

This document is not intended as an introduction to LISP rather it is
assumed that the reader is already familiar with some version.  The
document is thus intended as an arbiter of the syntax and semantics of
Standard LISP. However, since it is not intended as an implementation
description, we deliberately leave unspecified many of the details on
which an actual implementation depends. For example, while we assume
the existence of a symbol table for atoms (the "object list" in LISP
terminology), we do not specify its structure, since conventional LISP
programming does not require this information. Our ultimate goal,
however, is to remedy this by defining an interpreter for Standard
LISP which is sufficiently complete that its implementation on any
given computer will be straightforward and precise. At that time, we
shall produce an implementation level specification for Standard LISP
which will extend the description of the primitive functions defined
herein by introducing a new set of lower level primitive functions in
which the structure of the symbol table, heap and so on may be
defined.

The plan of this chapter is as follows. In Section~\ref{dtypes} we
describe the various data types used in Standard LISP. In
Section~\ref{slfns}, a description of all Standard LISP functions is
presented, organized by type. These functions are defined in an RLISP
syntax which is easier to read than LISP S-expressions.
Section~\ref{slglobals} describes global variables which control the
operation of Standard LISP.


\section{Preliminaries}
\label{dtypes}
\subsection{Primitive Data Types}
\label{pdat}
\begin{description}
\item[integer] Integers are also called "fixed" numbers. The magnitude of
an integer is unrestricted. Integers in the LISP input stream are
\index{integer ! input} \index{integer ! magnitude}
recognized by the grammar:

\begin{tabbing}
\s{digit} ::= 0$\mid$1$\mid$2$\mid$3$\mid$4$\mid$5$\mid$6$\mid$7$\mid$8$\mid$9
\\
\s{unsigned-integer} ::= \s{digit}$\mid$\s{unsigned-integer}\s{digit} \\
\s{integer} ::= \= \s{unsigned-integer} $\mid$ \\
\> +\s{unsigned-integer} $\mid$ \\
\> ---\s{unsigned-integer}
\end{tabbing}

\item[floating] - Any floating point number. The precision of floating point
\index{floating ! input}
numbers is determined solely by the implementation. In BNF floating
point numbers are recognized by the grammar:

\begin{tabbing}
\s{base} ::=  \= \s{unsigned-integer}.$\mid$.\s{unsigned-integer}$\mid$ \\
\> \s{unsigned-integer}.\s{unsigned-integer} \\
\> \s{unsigned-floating} ::= \s{base}$\mid$ \\
\> \s{base}E\s{unsigned-integer}$\mid$ \\
\> \s{base}E-\s{unsigned-integer}$\mid$ \\
\> \s{base}E+\s{unsigned-integer} \\
\s{floating} ::= \= \s{unsigned-floating}$\mid$ \\
\> +\s{unsigned-floating}$\mid$-\s{unsigned-floating}
\end{tabbing}

\item[id] An identifier is a string of characters which may have the
\index{id ! input} \index{identifier (see id)}
following items associated with it.

\begin{description}
\item[print name] \index{print name} The characters of the identifier.

\item[flags] An identifier may be tagged with a flag. Access is by the
FLAG, REMFLAG, and FLAGP functions defined in section~\ref{plist} on
page~\pageref{plist}. \index{FLAG} \index{REMFLAG} \index{FLAGP}

\item[properties] \index{properties} An identifier may have an
indicator-value pair associated with it. Access is by the PUT, GET,
and REMPROP functions defined in section~\ref{plist} on
page~\pageref{plist}.
\index{PUT} \index{GET} \index{REMPROP}

\item[values/functions] An identifier may have a value associated with
\index{values} \index{functions} it. Access to values is by SET and SETQ
defined in \index{SET} \index{SETQ} section~\ref{varsandbinds} on
page~\pageref{varsandbinds}. The method by which the value is attached
to the identifier is known as the binding type, being one of LOCAL,
GLOBAL, or FLUID. Access to the binding type is by the GLOBAL,
GLOBALP, FLUID, FLUIDP, and UNFLUID functions.
\index{GLOBAL} \index{GLOBALP} \index{FLUID} \index{FUIDP} \index{UNFLUID}

An identifier may have a function or macro associated with it. Access
is by the PUTD, GETD, and REMD functions (see ``Function Definition'',
section~\ref{fdef}, on page~\pageref{fdef}). \index{PUTD} \index{GETD}
\index{REMD} An identifier may not have both a function and a value
associated with it.

\item[OBLIST entry] \index{OBLIST entry} An identifier may be entered and
removed from a structure called the OBLIST. Its presence on the OBLIST
does not directly affect the other properties. Access to the OBLIST is
by the INTERN, REMOB, and READ functions. \index{INTERN} \index{REMOB}
\index{READ}
\end{description}

The maximum length of a Standard LISP identifier is 24 characters
\index{id ! maximum length}
(excluding occurrences of the escape character !) but an
\index{id ! escape character}
implementation may allow more. Special characters (digits in the first
position and punctuation) must be prefixed with an escape character,
an ! in Standard LISP. In BNF identifiers are recognized by the
grammar:


\begin{tabbing}
\s{special-character} ::= !\s{any-character} \\
\s{alphabetic} ::= \\
\hspace*{.25in} \= A$\mid$B$\mid$C$\mid$D$\mid$E$\mid$F$\mid$G$\mid$H$
\mid$I$\mid$J$\mid$K$\mid$L$\mid$M$\mid$N$\mid$O$\mid$P$\mid$Q$\mid$R$
\mid$S$\mid$T$\mid$U$\mid$V$\mid$W$\mid$X$\mid$Y$\mid$Z$\mid$ \\
\> a$\mid$b$\mid$c$\mid$d$\mid$e$\mid$f$\mid$g$\mid$h$\mid$i$\mid$j$
\mid$k$\mid$l$\mid$m$\mid$n$\mid$o$\mid$p$\mid$q$\mid$r$\mid$s$\mid$t$
\mid$u$\mid$v$\mid$w$\mid$x$\mid$y$\mid$z \\
\s{lead-character} ::= \s{special-character}$\mid$\s{alphabetic} \\
\s{regular-character} ::= \s{lead-character}$\mid$\s{digit} \\
\s{last-part} ::= \= \s{regular-character} $\mid$ \\
\> \s{last-part}\s{regular-character} \\
\s{id} ::= \s{lead-character}$\mid$\s{lead-character}\s{last-part}
\end{tabbing}

Note: Using lower case letters in identifiers may cause portability
problems. Lower case letters are automatically converted to upper case
when the !*RAISE flag is T. \index{*RAISE (global)}


\item[string] \index{string} A set of characters enclosed in double quotes as
in "THIS IS A STRING". A quote is included by doubling it as in "HE
SAID, ""LISP""". The maximum size of strings is 80 characters but an
implementation may allow more. Strings are not part of the OBLIST and
are considered constants like numbers, vectors, and function-pointers.

\item[dotted-pair] A primitive structure which has a left and right part.
\index{dotted-pair} \index{dot-notation}
A notation called {\em dot-notation} is used for dotted pairs and
takes the form:

\begin{tabbing}
(\s{left-part} . \s{right-part})
\end{tabbing}

The \s{left-part} is known as the CAR portion and the \s{right-part}
as the CDR portion. The left and right parts may be of any type.
Spaces are used to resolve ambiguity with floating point numbers.


\item[vector] \index{vector} A primitive uniform structure in which
an integer index is used to access random values in the structure. The
individual elements of a vector may be of any type. Access to vectors
is restricted to functions defined in ``Vectors''
section~\ref{vectors} on page~\pageref{vectors}. A notation for
vectors, {\em vector-notation}, has the elements of a vector
surrounded
\index{vector-notation}
by square brackets\footnote{Vector elements are not separated by
commas as in the published version of this document.}


\begin{tabbing}
\s{elements} ::= \s{any}$\mid$\s{any} \s{elements} \\
\s{vector} ::= [\s{elements}]
\end{tabbing}

\item[function-pointer] \index{function-pointer} An implementation may have
functions which deal with specific data types other than those listed.
The use of these entities is to be avoided with the exception of a
restricted use of the function-pointer, an access method to compiled
EXPRs and FEXPRs. A particular function-pointer must remain valid
\index{EXPR} \index{FEXPR}
throughout execution. Systems which change the location of a function
must use either an indirect reference or change all occurrences of the
associated value. There are two classes of use of function-pointers,
those which are supported by Standard LISP but are not well defined,
and those which are well defined.

\begin{description}
\item[Not well defined] Function pointers may be displayed by the print
functions or expanded by EXPLODE. \index{EXPLODE} The value appears in
the convention of the implementation site. The value is not defined in
Standard LISP. Function pointers may be created by COMPRESS
\index{COMPRESS} in the format used for printing but the value used is
not defined in Standard LISP. Function pointers may be created by
functions which deal with compiled function loading. Again, the values
created are not well defined in Standard LISP.

\item[Well defined] The function pointer associated with an EXPR or
FEXPR may be retrieved by GETD \index{GETD} and is valid as long as
Standard LISP is in execution. Function pointers may be stored using
\index{PUTD} \index{PUT} \index{SETQ} PUTD, PUT, SETQ and the like or by
being bound to variables.  Function pointers may be checked for
equivalence by EQ. \index{EQ ! of function-pointers} The value may be
checked for being a function pointer by the CODEP function.
\index{CODEP}
\end{description}
\end{description}


\subsection{Classes of Primitive Data Types}
\label{pclasses}
The classes of primitive types are a notational convenience for
describing the properties of functions.


\begin{description}
\item[boolean] \index{boolean} The set of global variables \{T,NIL\},
or their respective values, \{T, NIL\}. \index{T (global)} \index{NIL
(global)}

\item[extra-boolean] \index{extra-boolean} Any value in the system.
Anything that is not NIL \index{NIL (global)} has the boolean
interpretation T. \index{T (global)}

\item[ftype] \index{ftype} The class of definable function types. The
set of ids \{EXPR, FEXPR, MACRO\}. \index{EXPR} \index{FEXPR}
\index{MACRO}

\item[number] \index{number} The set of \{integer, floating\}.

\item[constant] \index{constant} The set of \{integer, floating,
string, vector, function-pointer\}. Constants evaluate to themselves
(see the definition of EVAL in ``The Interpreter'',
section~\ref{interpreter} on page~\pageref{interpreter}). \index{EVAL
! of constants}


\item[any] \index{any} The set of \{integer, floating, string, id,
dotted-pair, vector, function-pointer\}. An S-expression is another
term for any. All Standard LISP entities have some value unless an
ERROR occurs during evaluation or the function causes transfer of
control (such as GO and RETURN).


\item[atom] \index{atom} The set \{any\}-\{dotted-pair\}.
\end{description}

\subsection{Structures}
\index{data structures} \index{structures}
Structures are entities created out of the primitive types by the use
of dotted-pairs. Lists are structures very commonly required as actual
parameters to functions. Where a list of homogeneous entities is
required by a function this class will be denoted by
\s{{\bf xxx}-list} where {\bf \em xxx} is the name of a class of primitives
or structures. Thus a list of ids is an {\em id-list}, a list of
integers an {\em integer-list} and so on. \index{id-list}
\index{integer-list}
\index{-list}

\begin{description}
\item[list] \index{list} A list is recursively defined as NIL or the
\index{list-notation} \index{NIL (global)}
dotted-pair (any~.~list). A special notation called {\em
list-notation} is used to represent lists. List-notation eliminates
extra parentheses and dots. The list (a . (b . (c . NIL))) in list
notation is (a b c).
\index{dot-notation}
List-notation and dot-notation may be mixed as in (a b . c) or (a (b .
c) d) which are (a . (b . c)) and (a . ((b . c) . (d .  NIL))). In BNF
lists are recognized by the grammar:

\begin{tabbing}
\s{left-part} ::= ( $\mid$ \s{left-part} \s{any} \\
\s{list} ::= \s{left-part}) $\mid$ \s{left-part} . \s{any})
\end{tabbing}

Note: () is an alternate input representation of NIL. \index{()}


\item[alist] \index{alist} An association list; each element of the list
is a dotted-pair, the CAR part being a key associated with the value
in the CDR part. \index{association list}


\item[cond-form] \index{cond-form} A cond-form is a list of 2 element lists
of the form:

(\p{ANTECEDENT}:{\em any} \p{CONSEQUENT}:{\em any})

The first element will henceforth be known as the antecedent and
\index{antecedent (cond-form)} \index{consequent (cond-form)}
the second as the consequent. The antecedent must have a value.  The
consequent may have a value or an occurrence of GO or RETURN
\index{GO} \index{RETURN}
as described in the ``Program Feature Functions'', section~\ref{prog}
on page~\pageref{prog}.


\item[lambda] \index{LAMBDA} A LAMBDA expression which must have the form
(in list notation): (LAMBDA parameters body). ``parameters'' is a list
of formal parameters for ``body'' an S-expression to be evaluated. The
semantics of the evaluation are defined with the EVAL function (see
``The Interpreter'', section~\ref{interpreter} on \index{EVAL ! lambda
expressions} page~\pageref{interpreter}). \index{lambda expression}

\item[function] \index{function} A LAMBDA expression or a function-pointer
to a function. A function is always evaluated as an EVAL, SPREAD form.
\index{EVAL ! function}
\end{description} 


\subsection{Function Descriptions}

Each function is provided with a prototypical header line. Each formal
parameter is given a name and suffixed with its allowed type.  Lower
case, italic tokens are names of classes and upper case, bold face,
tokens are parameter names referred to in the definition. The type of
the value returned by the function (if any) is suffixed to the
parameter list.  If it is not commonly used the parameter type may be
a specific set enclosed in brackets \{\ldots\}. \index{\{\ldots\} ! as
syntax} For example:


\vspace{.1in}
\noindent \f{PUTD}(\p{FNAME}:\ty{id}, \p{TYPE}:\ty{ftype},
\p{BODY}:\{\ty{lambda, function-pointer}\}):\ty{id}
\vspace{.1in}

PUTD is a function with three parameters. The parameter FNAME is an id
to be the name of the function being defined. TYPE is the type of the
function being defined and BODY is a lambda expression or a
function-pointer. PUTD returns the name of the function being defined.



Functions which accept formal parameter lists of arbitrary length have
the type class and parameter enclosed in square brackets indicating
that zero or more occurrences of that argument are permitted.
\index{[\ldots] syntax} For example:

\vspace{.1in}
\noindent \f{AND}([\p{U}:\ty{any}]):\ty{extra-boolean}
\vspace{.1in}

AND is a function which accepts zero or more arguments which may be of
any type.

\subsection{Function Types}

EVAL type functions are those which are invoked with evaluated
\index{EVAL ! function type}
arguments. NOEVAL functions are invoked with unevaluated arguments.
\index{NOEVAL ! function type}
SPREAD type functions have their arguments passed in one-to-one
\index{SPREAD ! function type}
correspondence with their formal parameters. NOSPREAD functions
\index{NOSPREAD ! function type}
receive their arguments as a single list. EVAL, SPREAD functions are
\index{FEXPR}
associated with EXPRs and NO\-EVAL, NO\-SPREAD functions with FEXPRs.
EVAL, NO\-SPREAD and NOEVAL, SPREAD functions can be simulated using
NOEVAL, NO\-SPREAD functions or MACROs. \index{MACRO}

EVAL, SPREAD type functions may have a maximum of 15 parameters.
\index{formal parameter limit}
There is no limit on the number of parameters a NOEVAL, NOSPREAD
function or MACRO may have.

In the context of the description of an EVAL, SPREAD function, then we
speak of the formal parameters we mean their actual values.  However,
in a NOEVAL, NOSPREAD function it is the unevaluated actual
parameters.

A third function type, the MACRO, implements functions which
\index{MACRO}
create S-expressions based on actual parameters. When a macro
invocation is encountered, the body of the macro, a lambda expression,
is invoked as a NOEVAL, NOSPREAD function with the macro's invocation
bound as a list to the macros single formal parameter. When the macro
has been evaluated the resulting S-expression is reevaluated. The
description of the EVAL and EXPAND
\index{EVAL ! MACRO functions}
functions provide precise details.


\subsection{Error and Warning Messages}
\index{error messages}
Many functions detect errors. The description of such functions will
include these error conditions and suggested formats for display
\index{ERROR}
of the generated error messages. A call on the ERROR function is
implied but the error number is not specified by Standard LISP. In
some cases a warning message is sufficient. To distinguish between
\index{warning messages} \index{***** (error message)}
\index{*** (warning message)}
errors and warnings, errors are prefixed with five asterisks and
warnings with only three.

Primitive functions check arguments that must be of a certain
primitive type for being of that type and display an error message if
the argument is not correct. The type mismatch error always takes the
form:
\index{error ! type mismatch error}

\errormessage{***** PARAMETER not TYPE for FN}

Here PARAMETER is the unacceptable actual parameter, TYPE is the type
that PARAMETER was supposed to be. FN is the name of the function that
detected the error.

\subsection{Comments}

\index{comments} \index{\%}
The character \% signals the start of a comment, text to be ignored
during parsing.  A comment is terminated by the end of the line it
\index{READCH} \index{READ}
is on.  The function READCH must be able to read a comment one
character at a time.  Comments are transparent to the function READ.
\% may occur as a character in identifiers by preceding it with the
\index{escape character}
escape character !.


\section{Functions}
\label{slfns}

\subsection{Elementary Predicates}
\label{elpreds}
\index{predicate !}
\index{T (global)} \index{NIL (global)}
Functions in this section return T when the condition defined is met
and NIL when it is not. Defined are type checking functions and
elementary comparisons.


\de{ATOM}{(\p{U}:\ty{any}):{\ty boolean}}{eval, spread}
{Returns T if U is not a pair.

{\tt \begin{tabbing} EXPR PROCEDURE ATOM(U); \\
\hspace*{1em} NULL PAIRP U;
\end{tabbing}}}


\de{CODEP}{(\p{U}:\f{any}):{\ty boolean}}{eval, spread}
{Returns T if U is a function-pointer.}


\de{CONSTANTP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is a constant (a number, string, function-pointer, or
vector).

{\tt \begin{tabbing} EXPR PROCEDURE CONSTANTP(U); \\
\hspace*{1em} NULL OR(PAIRP U, IDP U);
\end{tabbing}}
}



\de{EQ}{(\p{U}:\ty{any}, \p{V}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U points to the same object as V. EQ is \underline{not}
a reliable comparison between numeric arguments. }


\de{EQN}{(\p{U}:\ty{any}, \p{V}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U and V are EQ or if U and V are numbers and have the
same value and type. }


\de{EQUAL}{(\p{U}:\ty{any}, \p{V}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U and V are the same. Dotted-pairs are compared
recursively to the bottom levels of their trees. Vectors must have
identical dimensions and EQUAL values in all positions. Strings must
\index{EQ ! of function-pointers} \index{EQN} have identical characters.
Function pointers must have EQ values. Other atoms must be EQN equal. }


\de{FIXP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is an integer (a fixed number).}


\de{FLOATP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is a floating point number. }


\de{IDP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is an id.}


\de{MINUSP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is a number and less than 0.  If U is not a number or
is a positive number, NIL is returned.

{\tt \begin{tabbing} EXPR PROCEDURE MINUSP(U); \\
\hspace*{1em} IF NUMBERP U THEN LESSP(U, 0) ELSE NIL;
\end{tabbing}}}


\de{NULL}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is NIL.

{\tt \begin{tabbing} EXPR PROCEDURE NULL(U); \\
\hspace*{1em} U EQ NIL;
\end{tabbing}}}


\de{NUMBERP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is a number (integer or floating).

{\tt \begin{tabbing} EXPR PROCEDURE NUMBERP(U); \\
\hspace*{1em} IF OR(FIXP U, FLOATP U) THEN T ELSE NIL;
\end{tabbing}}}


\de{ONEP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread.}
{Returns T if U is a number and has the value 1 or 1.0.  Returns NIL
otherwise. \footnote{The definition in the published report is
incorrect as it does not return T for \p{U} of 1.0.}

{\tt \begin{tabbing} EXPR PROCEDURE ONEP(U); \\
\hspace*{1em} OR(EQN(U, 1), EQN(U, 1.0));
\end{tabbing}}}


\de{PAIRP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is a dotted-pair. }


\de{STRINGP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is a string. }


\de{VECTORP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is a vector. }


\de{ZEROP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread.}
{Returns T if U is a number and has the value 0 or 0.0.  Returns NIL
otherwise.\footnote{The definition in the published report is
incorrect as it does not return T for \p{U} of 0.0.}

{\tt \begin{tabbing} EXPR PROCEDURE ZEROP(U); \\
\hspace*{1em} OR(EQN(U, 0), EQN(U, 0.0));
\end{tabbing}}}


\subsection{Functions on Dotted-Pairs}

\index{dotted-pair}
The following are elementary functions on dotted-pairs. All functions
in this section which require dotted-pairs as parameters detect a type
mismatch error if the actual parameter is not a dotted-pair.



\de{CAR}{(\p{U}:\ty{dotted-pair}):\ty{any}}{eval, spread}
{CAR(CONS(a, b)) $\rightarrow$ a. The left part of U is returned. The
type
\index{CONS}
mismatch error occurs if U is not a dotted-pair.}


\de{CDR}{(\p{U}:\ty{dotted-pair}):\ty{any}}{eval, spread}
{CDR(CONS(a, b)) $\rightarrow$ b. The right part of U is returned. The
type
\index{CONS}
mismatch error occurs if U is not a dotted-pair.}


The composites of CAR and CDR are supported up to 4 levels, namely:
\index{CAR ! composite forms} \index{CDR ! composite forms}

\hspace*{1in}\begin{tabular}{l l l}
CAAAAR & CAAAR & CAAR \\ CAAADR & CAADR & CADR \\ CAADAR & CADAR &
CDAR \\ CAADDR & CADDR & CDDR \\ CADAAR & CDAAR & \\ CADADR & CDADR &
\\ CADDAR & CDDAR & \\ CADDDR & CDDDR & \\ CDAAAR & & \\ CDAADR & & \\
CDADAR & & \\ CDADDR & & \\ CDDAAR & & \\ CDDADR & & \\ CDDDAR & & \\
CDDDDR & &
\end{tabular}

\de{CONS}{(\p{U}:\ty{any}, \p{V}:\ty{any}):\ty{dotted-pair}}{eval, spread}
{Returns a dotted-pair which is not EQ to anything and has U as its
\index{EQ ! of dotted-pairs} \index{dotted-pair}
CAR part and V as its CDR part.}


\de{LIST}{([\p{U}:\ty{any}]):\ty{list}}{noeval, nospread, or macro}
{A list of the evaluation of each element of U is returned. The order
of evaluation need not be first to last as the following definition
implies.\footnote{The published report's definition implies a specific
ordering.}

{\tt \begin{tabbing} FEXPR PROCEDURE LIST(U); \\
\hspace*{1em} EVLIS U;
\end{tabbing}}}


\de{RPLACA}{(\p{U}:\ty{dotted-pair},
\p{V}:\ty{any}):\ty{dotted-pair}}{eval, spread}
{The CAR portion of the dotted-pair U is replaced by V. If dotted-pair
U is (a . b) then (V . b) is returned. The type mismatch error occurs
if U is not a dotted-pair. }


\de{RPLACD}{(\p{U}:\ty{dotted-pair},
\p{V}:\ty{any}):\ty{dotted-pair}}{eval, spread}
{The CDR portion of the dotted-pair U is replaced by V. If dotted-pair
U is (a . b) then (a . V) is returned. The type mismatch error occurs
if U is not a dotted-pair.}


\subsection{Identifiers}
\label{identifiers}
The following functions deal with identifiers and the OBLIST,
\index{OBLIST}
the structure of which is not defined. The function of the OBLIST is
to provide a symbol table for identifiers created during input.
Identifiers created by READ which have the same characters will
\index{READ} \index{EQ ! of identifiers}
therefore refer to the same object (see the EQ function in
``Elementary Predicates'', section~\ref{elpreds} on
page~\pageref{elpreds}).



\de{COMPRESS}{(\p{U}:\ty{id-list}):\{\ty{atom}-\ty{vector}\}}{eval, spread}
{U is a list of single character identifiers which is built into a
Standard LISP entity and returned. Recognized are numbers, strings,
and identifiers with the escape character prefixing special
characters. The formats of these items appear in ``Primitive Data
Types'' section~\ref{pdat} on page~\pageref{pdat}. Identifiers are not
interned on the OBLIST. Function pointers may be compressed but this
is an undefined use. If an entity cannot be parsed out of U or
characters are left over after parsing an error occurs:

\errormessage{***** Poorly formed atom in COMPRESS}
}


\de{EXPLODE}{(\p{U}:\{\ty{atom}\}-\{\ty{vector}\}):\ty{id-list}}{eval, spread}
{Returned is a list of interned characters representing the characters
to print of the value of U. The primitive data types have these
formats:

\begin{description}
\item[integer] \index{integer ! output} Leading zeroes are suppressed and
a minus sign prefixes the digits if the integer is negative.

\item[floating] \index{floating ! output} The value appears in the format
[-]0.nn...nnE[-]mm if the magnitude of the number is too large or
small to display in [-]nnnn.nnnn format. The crossover point is
determined by the implementation.

\item[id] \index{id ! output} The characters of the print name of the
identifier are produced with special characters prefixed with the
escape character.

\item[string] \index{string ! output} The characters of the string are
produced surrounded by double quotes "\ldots".

\item[function-pointer] \index{function-pointer ! output} The value of the
function-pointer is created as a list of characters conforming to the
conventions of the system site.
\end{description}

The type mismatch error occurs if U is not a number, identifier,
string, or function-pointer. }


\de{GENSYM}{():\ty{identifier}}{eval, spread}
{Creates an identifier which is not interned on the OBLIST and
consequently not EQ to anything else. \index{OBLIST entry} \index{EQ !
of GENSYMs}}


\de{INTERN}{(\p{U}:\{\ty{id,string}\}):\ty{id}}{eval, spread}
{INTERN searches the OBLIST for an identifier with the same print name
\index{OBLIST entry}
as U and returns the identifier on the OBLIST if a match is found.
Any properties and global values associated with U may be lost. If U
does not match any entry, a new one is created and returned. If U has
more than the maximum number of characters permitted by the
implementation (the minimum number is 24) an error occurs:
\index{id ! minimum size}

\errormessage{***** Too many characters to INTERN}
}


\de{REMOB}{(\p{U}:\ty{id}):\ty{id}}{eval, spread}
{If U is present on the OBLIST it is removed. This does not affect U
\index{OBLIST entry}
having properties, flags, functions and the like. U is returned.}


\subsection{Property List Functions}
\label{plist}
\index{property list}
With each id in the system is a ``property list'', a set of entities
which are associated with the id for fast access. These entities are
called ``flags'' if their use gives the id a single valued
\index{flags}
property, and ``properties'' if the id is to have a multivalued
\index{properties}
attribute: an indicator with a property.

Flags and indicators may clash, consequently care should be taken to
avoid this occurrence. Flagging X with an id which already is an
indicator for X may result in that indicator and associated property
being lost. Likewise, adding an indicator which is the same id as a
flag may result in the flag being destroyed.



\de{FLAG}{(\p{U}:\ty{id-list}, \p{V}:\ty{id}):\ty{NIL}}{eval, spread}
{U is a list of ids which are flagged with V. The effect of FLAG is
that FLAGP will have the value T for those ids of U which were
flagged. Both V and all the elements of U must be identifiers or the
type mismatch error occurs.}


\de{FLAGP}{(\p{U}:\ty{any}, \p{V}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U has been previously flagged with V, else NIL. Returns
NIL if either U or V is not an id.}


\de{GET}{(\p{U}:\ty{any}, \p{IND}:\ty{any}):\ty{any}}{eval, spread}
{Returns the property associated with indicator IND from the property
list of U. If U does not have indicator IND, NIL is returned. GET
cannot be used to access functions (use GETD instead).
\index{GET ! not for functions}}


\de{PUT}{(\p{U}:\ty{id}, \p{IND}:\ty{id},
\p{PROP}:\ty{any}):\ty{any}}{eval, spread}
{The indicator IND with the property PROP is placed on the property
list of the id U. If the action of PUT occurs, the value of PROP is
returned. If either of U and IND are not ids the type mismatch error
will occur and no property will be placed. PUT cannot be used to
define functions (use PUTD instead).
\index{PUT ! not for functions}}


\de{REMFLAG}{(\p{U}:\ty{any-list}, \p{V}:\ty{id}):\ty{NIL}}{eval, spread}
{Removes the flag V from the property list of each member of the list
U. Both V and all the elements of U must be ids or the type mismatch
error will occur.}


\de{REMPROP}{(\p{U}:\ty{any}, \p{IND}:\ty{any}):\ty{any}}{eval, spread}
{Removes the property with indicator IND from the property list of U.
Returns the removed property or NIL if there was no such indicator.}



\subsection{Function Definition}
\label{fdef}
Functions in Standard LISP are global entities. To avoid
function-variable naming clashes no variable may have the same name as
a function. \index{function ! as GLOBAL}


\de{DE}{(\p{FNAME}:\ty{id}, \p{PARAMS}:\ty{id-list},
\p{FN}:\ty{any}):\ty{id}}{noeval, nospread}
{The function FN with the formal parameter list PARAMS is added to the
set of defined functions with the name FNAME. Any previous definitions
of the function are lost. The function created is of type
\index{*COMP (fluid)} 
EXPR.  If the !*COMP variable is non-NIL, the EXPR is first
\index{EXPR}
compiled. The name of the defined function is returned.

{\tt \begin{tabbing} FEXPR PROCEDURE DE(U); \\
\hspace*{1em} PUTD(CAR U, 'EXPR, LIST('LAMBDA, CADR U, CADDR U));
\end{tabbing}}}


\de{DF}{(\p{FNAME}:\ty{id}, \p{PARAM}:\ty{id-list},
\p{FN}:\ty{any}):\ty{id}}{noeval, nospread}
{The function FN with formal parameter PARAM is added to the set of
defined functions with the name FNAME. Any previous definitions of the
function are lost. The function created is of type FEXPR.
\index{*COMP variable} \index{FEXPR}
If the !*COMP variable is T the FEXPR is first compiled. The name of
the defined function is returned.

{\tt \begin{tabbing} FEXPR PROCEDURE DF(U); \\
\hspace*{1em} PUTD(CAR U, 'FEXPR, LIST('LAMBDA, CADR U, CADDR U)); \\
\end{tabbing} }}


\de{DM}{(\p{MNAME}:\ty{id}, \p{PARAM}:\ty{id-list},
\p{FN}:\ty{any}):\ty{id}}{noeval, nospread}
{The macro FN with the formal parameter PARAM is added to the set of
defined functions with the name MNAME. Any previous definitions of the
function are overwritten. The function created is of type MACRO.
\index{MACRO}
The name of the macro is returned.

{\tt \begin{tabbing} FEXPR PROCEDURE DM(U); \\
\hspace*{1em} PUTD(CAR U, 'MACRO, LIST('LAMBDA, CADR U, CADDR U));
\end{tabbing} }
}


\de{GETD}{(\p{FNAME}:\ty{any}):\{NIL, \ty{dotted-pair}\}}{eval, spread}
{If FNAME is not the name of a defined function, NIL is returned. If
FNAME is a defined function then the dotted-pair

\vspace{.15in}
(\p{TYPE}:\ty{ftype} . \p{DEF}:\{\ty{function-pointer, lambda}\})
\vspace{.15in}

is returned.}


\de{PUTD}{(\p{FNAME}:\ty{id}, \p{TYPE}:\ty{ftype},
\p{BODY}:\ty{function}):\ty{id}}{eval, spread}
{Creates a function with name FNAME and definition BODY of type TYPE.
If PUTD succeeds the name of the defined function is returned. The
effect of PUTD is that GETD will return a dotted-pair with the
functions type and definition. Likewise the GLOBALP predicate will
\index{GLOBALP} \index{function ! as global}
return T when queried with the function name.

If the function FNAME has already been declared as a GLOBAL or FLUID
variable the error:

\errormessage{***** FNAME is a non-local variable}

occurs and the function will not be defined. If function FNAME already
exists a warning message will appear:

\errormessage{*** FNAME redefined}

The function defined by PUTD will be compiled before definition
\index{*COMP (fluid)} if the !*COMP global variable is non-NIL.}


\de{REMD}{(\p{FNAME}:\ty{id}):\{NIL, \ty{dotted-pair}\}}{eval, spread}
{Removes the function named FNAME from the set of defined functions.
Returns the (ftype . function) dotted-pair or NIL as does GETD. The
global/function attribute of FNAME is removed and the name may be used
subsequently as a variable.}



\subsection{Variables and Bindings}
\label{varsandbinds}
\index{variable scope} \index{scope}
A variable is a place holder for a Standard LISP entity which is said
to be bound to the variable. The scope of a variable is the range over
which the variable has a defined value. There are three different
binding mechanisms in Standard LISP.

\begin{description}
\item[Local Binding] \index{local binding} This type of binding occurs
\index{scope ! local}
only in compiled functions. Local variables occur as formal parameters
in lambda expressions and as PROG form variables. The binding occurs
when a lambda expression is evaluated or when a PROG form is executed.
The scope of a local variable is the body of the function in which it
is defined.

\item[Global Binding] \index{global binding} Only one binding of a
\index{scope ! global}
global variable exists at any time allowing direct access to the value
bound to the variable.  The scope of a global variable is universal.
Variables declared GLOBAL may not appear as parameters in lambda
expressions or as PROG form variables. A variable must be declared
GLOBAL prior to its use as a global variable since the default type
for undeclared variables is FLUID.


\item[Fluid Binding] \index{fluid binding}
\index{fluid binding ! as default} Fluid variables are global
in scope but may occur as \index{scope ! fluid} formal parameters or
PROG form variables. In interpreted functions all formal parameters
and PROG form variables are considered to have fluid binding until
changed to local binding by compilation.  When fluid variables are
used as parameters they are rebound in such a way that the previous
binding may be restored. All references to fluid variables are to the
currently active binding.
\end{description}


\de{FLUID}{(\p{IDLIST}:\ty{id-list}):\p{NIL}}{eval, spread}
{The ids in IDLIST are declared as FLUID type variables (ids not
previously declared are initialized to NIL). Variables in IDLIST
already declared FLUID are ignored. Changing a variable's type from
GLOBAL to FLUID is not permissible and results in the error:

\errormessage{***** ID cannot be changed to FLUID}
}

\de{FLUIDP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{If U has been declared FLUID (by declaration only) T is returned,
otherwise NIL is returned.}


\de{GLOBAL}{(\p{IDLIST}:\ty{id-list}):\p{NIL}}{eval, spread}
{The ids of IDLIST are declared global type variables. If an id has
not been declared previously it is initialized to NIL. Variables
already declared GLOBAL are ignored. Changing a variables type from
FLUID to GLOBAL is not permissible and results in the error:

\errormessage{***** ID cannot be changed to GLOBAL}
}


\de{GLOBALP}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{If U has been declared GLOBAL or is the name of a defined function, T
is returned, else NIL is returned.}


\de{SET}{(\p{EXP}:\ty{id}, \p{VALUE}:\ty{any}):\ty{any}}{eval, spread}
{EXP must be an identifier or a type mismatch error occurs. The effect
of SET is replacement of the item bound to the identifier by VALUE.
If the identifier is not a local variable or has not been declared
GLOBAL it is automatically declared FLUID with the resulting warning
message:

\errormessage{*** EXP declared FLUID}

EXP must not evaluate to T or NIL or an error occurs:
\index{T ! cannot be changed} \index{NIL ! cannot be changed}

\errormessage{***** Cannot change T or NIL}
}

\de{SETQ}{(\p{VARIABLE}:\ty{id}, \p{VALUE}:\ty{any}):\ty{any}}{noeval,
nospread}
{If VARIABLE is not local or GLOBAL it is by default declared FLUID
and the warning message:

\errormessage{*** VARIABLE declared FLUID}

appears. The value of the current binding of VARIABLE is replaced by
the value of VALUE. VARIABLE must not be T or NIL or an error occurs:
\index{T ! cannot be changed} \index{NIL ! cannot be changed}

\errormessage{***** Cannot change T or NIL}

{\tt \begin{tabbing} MACRO PROCEDURE SETQ(X); \\
\hspace*{1em} LIST('SET, LIST('QUOTE, CADR X), CADDR X);
\end{tabbing}}
}

\de{UNFLUID}{(\p{IDLIST}:\ty{id-list}):\ty{NIL}}{eval, spread}
{The variables in IDLIST that have been declared as FLUID variables
are no longer considered as fluid variables. Others are ignored. This
affects only compiled functions as free variables in interpreted
functions are automatically considered fluid~\cite{PLC}.
\index{scope ! fluid and compiled}}


\subsection{Program Feature Functions}
\label{prog}
These functions provide for explicit control sequencing, and the
definition of blocks altering the scope of local variables.


\de{GO}{(\p{LABEL}:\ty{id})}{noeval, nospread}
{GO alters the normal flow of control within a PROG function. The next
statement of a PROG function to be evaluated is immediately preceded
by LABEL. A GO may only appear in the following situations:


\begin{enumerate}
\item At the top level of a PROG referencing a label which also
appears at the top level of the same PROG.

\item As the consequent of a COND item of a COND appearing on the top
level of a PROG.
\index{GO ! in COND}
\index{RETURN ! in COND}
\item As the consequent of a COND item which appears as the
consequent of a COND item to any level.

\item As the last statement of a PROGN which appears at the top level
of a PROG or in a PROGN appearing in the consequent of a COND to any
level subject to the restrictions of 2 and 3.

\item As the last statement of a PROGN within a PROGN or as the
consequent of a COND in a PROGN to any level subject to the
restrictions of 2, 3 and 4.
\end{enumerate}

If LABEL does not appear at the top level of the PROG in which the GO
appears, an error occurs:

\errormessage{***** LABEL is not a known label}

If the GO has been placed in a position not defined by rules 1-5,
another error is detected:

\errormessage{***** Illegal use of GO to LABEL}
}

\de{PROG}{(\p{VARS}:\ty{id-list},
[\p{PROGRAM}:\{\ty{id, any}\}]):\ty{any}}{noeval, nospread}
{VARS is a list of ids which are considered fluid when the PROG is
interpreted and local when compiled (see ``Variables and Bindings'',
section~\ref{varsandbinds} on page~\pageref{varsandbinds}). The PROGs
variables are allocated space when the PROG form is invoked and are
deallocated when the PROG is exited. PROG variables are initialized to
\index{PROG ! variables}
NIL. The PROGRAM is a set of expressions to be evaluated in order of
their appearance in the PROG function.  Identifiers appearing in the
top level of the PROGRAM are labels which can be referenced by GO. The
value returned by the PROG function is determined by a RETURN function
\index{PROG ! default value}
or NIL if the PROG ``falls through''.}


\de{PROGN}{([\p{U}:\ty{any}]):\ty{any}}{noeval, nospread}
{U is a set of expressions which are executed sequentially. The value
returned is the value of the last expression.}


\de{PROG2}{(A:any, B:any)\ty{any}}{eval, spread}
{Returns the value of B.

{\tt \begin{tabbing} EXPR PROCEDURE PROG2(A, B);\\
\hspace*{1em} B;
\end{tabbing}}}


\de{RETURN}{(\p{U}:\ty{any})}{eval, spread}
{Within a PROG, RETURN terminates the evaluation of a PROG and returns
U as the value of the PROG. The restrictions on the placement of
RETURN are exactly those of GO. Improper placement of RETURN results
in the error:

\errormessage{***** Illegal use of RETURN}
}


\subsection{Error Handling}
\label{errors}

\de{ERROR}{(\p{NUMBER}:\ty{integer}, \p{MESSAGE}:\ty{any})}{eval, spread}
{NUMBER and MESSAGE are passed back to a surrounding ERRORSET (the
Standard LISP reader has an ERRORSET). MESSAGE is placed in the
\index{EMSG* (global)}
global variable EMSG!* and the error number becomes the value of the
surrounding ERRORSET. FLUID variables and local bindings are unbound
\index{fluid ! unbinding by ERROR}
to return to the environment of the ERRORSET. Global variables are not
affected by the process.}


\de{ERRORSET}{(\p{U}:\ty{any}, \p{MSGP}:\ty{boolean},
\p{TR}:\ty{boolean}):\ty{any}}{eval, spread}
{If an error occurs during the evaluation of U, the value of NUMBER
from the ERROR call is returned as the value of ERRORSET. In addition,
if the value of MSGP is non-NIL, the MESSAGE from the ERROR call is
displayed upon both the standard output device and the currently
selected output device unless the standard output device is not open.
The message appears prefixed with 5 asterisks. The MESSAGE
\index{***** (error message)}
list is displayed without top level parentheses. The MESSAGE from the
\index{EMSG* (global)}
ERROR call will be available in the global variable EMSG!*. The exact
format of error messages generated by Standard LISP functions
described in this document are not fixed and should not be relied upon
to be in any particular form. Likewise, error numbers generated by
Standard LISP functions are implementation dependent.

If no error occurs during the evaluation of U, the value of (LIST
(EVAL U)) is returned.

If an error has been signaled and the value of TR is non-NIL a
traceback sequence will be initiated on the selected output device.
The traceback will display information such as unbindings of FLUID
\index{fluid ! in traceback}
variables, argument lists and so on in an implementation dependent
format.}


\subsection{Vectors}
\label{vectors}
\index{vector}
Vectors are structured entities in which random elements may be
accessed with an integer index. A vector has a single dimension. Its
maximum size is determined by the implementation and available space.
A suggested input ``vector notation'' is defined in ``Classes of
Primitive Data Types'', section~\ref{pclasses} on
page~\pageref{pclasses} and output with EXPLODE, ``Identifiers''
section~\ref{identifiers} on page~\pageref{identifiers}.
\index{EXPLODE}


\de{GETV}{(\p{V}:\ty{vector}, \p{INDEX}:\ty{integer}):\ty{any}}{eval, spread}
{Returns the value stored at position INDEX of the vector V. The type
mismatch error may occur. An error occurs if the INDEX does not lie
within 0\ldots UPBV(V) inclusive:

\errormessage{***** INDEX subscript is out of range}
}


\de{MKVECT}{(\p{UPLIM}:\ty{integer}):\ty{vector}}{eval, spread}
{Defines and allocates space for a vector with UPLIM+1 elements
accessed as 0\ldots UPLIM. Each element is initialized to NIL. An
error will occur if UPLIM is $<$ 0 or there is not enough space for a
vector of this size:

\errormessage{***** A vector of size UPLIM cannot be allocated}
}


\de{PUTV}{(\p{V}:\ty{vector}, \p{INDEX}:\ty{integer},
\p{VALUE}:\ty{any}):\ty{any}}{eval, spread}
{Stores VALUE into the vector V at position INDEX. VALUE is returned.
The type mismatch error may occur. If INDEX does not lie in 0\ldots
UPBV(V) an error occurs:

\errormessage{***** INDEX subscript is out of range}
}


\de{UPBV}{(\p{U}:\ty{any}):{NIL,\ty{integer}}}{eval, spread}
{Returns the upper limit of U if U is a vector, or NIL if it is not.}


\subsection{Boolean Functions and Conditionals}


\de{AND}{([\p{U}:\ty{any}]):\ty{extra-boolean}}{noeval, nospread}
{AND evaluates each U until a value of NIL is found or the end of the
list is encountered. If a non-NIL value is the last value it is
returned, or NIL is returned.

{\tt \begin{tabbing} FEXPR PROCEDURE AND(U); \\ BEGIN \\
\hspace*{1em} IF NULL U THEN RETURN NIL; \\
LOOP: IF \= NULL CDR U THEN RETURN EVAL CAR U \\
\> ELSE IF NULL EVAL CAR U THEN RETURN NIL; \\
\hspace*{2em} \= U := CDR U; \\
\> GO LOOP \\
END;
\end{tabbing} }}


\de{COND}{([\p{U}:\ty{cond-form}]):\ty{any}}{noeval, nospread}
{The antecedents of all U's are evaluated in order of their appearance
until a non-NIL value is encountered. The consequent of the selected U
is evaluated and becomes the value of the COND. The consequent may
also contain the special functions GO and RETURN subject to the
restraints given for these functions in ``Program Feature Functions'',
section~\ref{prog} on page~\pageref{prog}.
\index{GO ! in COND} \index{RETUNR ! in CODE} In these cases COND does
not have a defined value, but rather an effect. If no antecedent is
non-NIL the value of COND is NIL. An error is detected if a U is
improperly formed:

\errormessage{***** Improper cond-form as argument of COND}
}


\de{NOT}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{If U is NIL, return T else return NIL (same as function NULL).

{\tt \begin{tabbing} EXPR PROCEDURE NOT(U); \\
\hspace*{1em} U EQ NIL;
\end{tabbing}}
}


\de{OR}{([\p{U}:\ty{any}]):\ty{extra-boolean}}{noeval, nospread}
{U is any number of expressions which are evaluated in order of their
appearance. When one is found to be non-NIL it is returned as the
value of OR. If all are NIL, NIL is returned.

{\tt \begin{tabbing} FEXPR PROCEDURE OR(U); \\ BEGIN SCALAR X; \\
LOOP: IF \= NULL U THEN RETURN NIL \\
\> ELSE IF (X := EVAL CAR U) THEN RETURN X; \\
\hspace*{2em} \= U := CDR U; \\
\> GO LOOP \\
END;
\end{tabbing} }}


\subsection{Arithmetic Functions}

Conversions between numeric types are provided explicitly by the
\index{FIX} \index{FLOAT}
FIX and FLOAT functions and implicitly by any multi-parameter
\index{mixed-mode arithmetic}
arithmetic function which receives mixed types of arguments. A
conversion from fixed to floating point numbers may result in a loss
of precision without a warning message being generated. Since
\index{integer ! magnitude}
integers may have a greater magnitude that that permitted for floating
numbers, an error may be signaled when the attempted conversion cannot
be done. Because the magnitude of integers is unlimited the conversion
of a floating point number to a fixed number is always possible, the
only loss of precision being the digits to the right of the decimal
point which are truncated. If a function receives mixed types of
arguments the general rule will have the fixed numbers converted to
floating before arithmetic operations are performed. In all cases an
error occurs if the parameter to an arithmetic function is not a
number:

\errormessage{***** XXX parameter to FUNCTION is not a number}

XXX is the value of the parameter at fault and FUNCTION is the name of
the function that detected the error. Exceptions to the rule are noted
where they occur.




\de{ABS}{(\p{U}:\ty{number}):\ty{number}}{eval, spread}
{Returns the absolute value of its argument.

{\tt \begin{tabbing} EXPR PROCEDURE ABS(U); \\
\hspace*{1em} IF LESSP(U, 0) THEN MINUS(U) ELSE U;
\end{tabbing}}}

\de{ADD1}{(\p{U}:\ty{number}):\ty{number}}{eval, spread}
{Returns the value of U plus 1 of the same type as U (fixed or
floating).

{\tt \begin{tabbing} EXPR PROCEDURE ADD1(U); \\
% God knows why, but hspace* isn't accepted here.
\hspace{1em} PLUS2(U, 1);
\end{tabbing}}
}

\de{DIFFERENCE}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{number}}{eval,
spread}
{The value U - V is returned.}


\de{DIVIDE}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{dotted-pair}}{eval,
spread}
{The dotted-pair (quotient . remainder) is returned. The quotient part
is computed the same as by QUOTIENT and the remainder the same as by
REMAINDER. An error occurs if division by zero is attempted:
\index{division by zero}

\errormessage{***** Attempt to divide by 0 in DIVIDE}

{\tt \begin{tabbing} EXPR PROCEDURE DIVIDE(U, V); \\
\hspace*{1em} (QUOTIENT(U, V) . REMAINDER(U, V));
\end{tabbing}}}


\de{EXPT}{(\p{U}:\ty{number}, \p{V}:\ty{integer}):\ty{number}}{eval, spread}
{Returns U raised to the V power. A floating point U to an integer
power V does \underline{not} have V changed to a floating number
before exponentiation.}


\de{FIX}{(\p{U}:\ty{number}):\ty{integer}}{eval, spread}
{Returns an integer which corresponds to the truncated value of U. The
result of conversion must retain all significant portions of U. If U
is an integer it is returned unchanged. }


\de{FLOAT}{(\p{U}:\ty{number}):\ty{floating}}{eval, spread}
{The floating point number corresponding to the value of the argument
U is returned. Some of the least significant digits of an integer may
be lost do to the implementation of floating point numbers. FLOAT of a
floating point number returns the number unchanged. If U is too large
to represent in floating point an error occurs:

\errormessage{***** Argument to FLOAT is too large}
}

\de{GREATERP}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{boolean}}{eval,
spread}
{Returns T if U is strictly greater than V, otherwise returns NIL.}


\de{LESSP}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{boolean}}{eval, spread}
{Returns T if U is strictly less than V, otherwise returns NIL. }


\de{MAX}{([\p{U}:\ty{number}]):\ty{number}}{noeval, nospread, or macro}
{Returns the largest of the values in U. If two or more values are the
same the first is returned.

{\tt \begin{tabbing} MACRO PROCEDURE MAX(U); \\
\hspace*{1em} EXPAND(CDR U, 'MAX2);
\end{tabbing}}}


\de{MAX2}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{number}}{eval, spread}
{Returns the larger of U and V. If U and V are the same value U is
returned (U and V might be of different types).

{\tt \begin{tabbing} EXPR PROCEDURE MAX2(U, V); \\
\hspace*{1em} IF LESSP(U, V) THEN V ELSE U;
\end{tabbing}}}


\de{MIN}{([\p{U}:\ty{number}]):\ty{number}}{noeval, nospread, or macro}
{Returns the smallest of the values in U. If two or more values are
the same the first of these is returned.

{\tt \begin{tabbing} MACRO PROCEDURE MIN(U); \\
\hspace*{1em} EXPAND(CDR U, 'MIN2);
\end{tabbing}}}


\de{MIN2}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{number}}{eval, spread}
{Returns the smaller of its arguments. If U and V are the same value,
U is returned (U and V might be of different types).

{\tt \begin{tabbing} EXPR PROCEDURE MIN2(U, V); \\
\hspace*{1em} IF GREATERP(U, V) THEN V ELSE U;
\end{tabbing}}}


\de{MINUS}{(\p{U}:\ty{number}):\ty{number}}{eval, spread}
{Returns -U.

{\tt \begin{tabbing} EXPR PROCEDURE MINUS(U); \\
\hspace*{1em} DIFFERENCE(0, U);
\end{tabbing}}}


\de{PLUS}{([\p{U}:\ty{number}]):\ty{number}}{noeval, nospread, or macro}
{Forms the sum of all its arguments.

{\tt \begin{tabbing} MACRO PROCEDURE PLUS(U); \\
\hspace*{1em} EXPAND(CDR U, 'PLUS2);
\end{tabbing}}}

\de{PLUS2}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{number}}{eval, spread}
{Returns the sum of U and V.}


\de{QUOTIENT}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{number}}{eval, spread}
{The quotient of U divided by V is returned. Division of two positive
or two negative integers is conventional. When both U and V are
integers and exactly one of them is negative the value returned is the
negative truncation of the absolute value of U divided by the absolute
value of V. An error occurs if division by zero is attempted:
\index{division by zero}

\errormessage{***** Attempt to divide by 0 in QUOTIENT}
}

\de{REMAINDER}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{number}}{eval,
spread}
{If both U and V are integers the result is the integer remainder of U
divided by V. If either parameter is floating point, the result is the
difference between U and V*(U/V) all in floating point. If either
number is negative the remainder is negative. If both are positive or
both are negative the remainder is positive. An error occurs if V is
zero: \index{division by zero}

\errormessage{***** Attempt to divide by 0 in REMAINDER}

{\tt \begin{tabbing} EXPR PROCEDURE REMAINDER(U, V); \\
\hspace*{1em} DIFFERENCE(U, TIMES2(QUOTIENT(U, V), V));
\end{tabbing}}}


\de{SUB1}{(\p{U}:\ty{number}):\ty{number}}{eval, spread}
{Returns the value of U less 1.  If U is a FLOAT type number, the
value returned is U less 1.0.

{\tt \begin{tabbing} EXPR PROCEDURE SUB1(U); \\
\hspace*{1em} DIFFERENCE(U, 1);
\end{tabbing}}}


\de{TIMES}{([\p{U}:\ty{number}]):\ty{number}}{noeval, nospread, or macro}
{Returns the product of all its arguments.

{\tt \begin{tabbing} MACRO PROCEDURE TIMES(U); \\
\hspace*{1em} EXPAND(CDR U, 'TIMES2);
\end{tabbing}}}


\de{TIMES2}{(\p{U}:\ty{number}, \p{V}:\ty{number}):\ty{number}}{eval, spread}
{Returns the product of U and V.}


\subsection{MAP Composite Functions}


\de{MAP}{(\p{X}:\ty{list}, F\p{N}:\ty{function}):\ty{any}}{eval, spread}
{Applies FN to successive CDR segments of X. NIL is returned.

{\tt \begin{tabbing} EXPR PROCEDURE MAP(X, FN); \\
\hspace*{1em} WHILE X DO $<<$ FN X; X := CDR X $>>$;
\end{tabbing}}}


\de{MAPC}{(X:list, FN:function):\ty{any}}{eval, spread}
{FN is applied to successive CAR segments of list X. NIL is returned.

{\tt \begin{tabbing} EXPR PROCEDURE MAPC(X, FN); \\
\hspace*{1em} WHILE X DO $<<$ FN CAR X; X := CDR X $>>$;
\end{tabbing}}}


\de{MAPCAN}{(X:list, FN:function):\ty{any}}{eval, spread}
{A concatenated list of FN applied to successive CAR elements of X is
returned.

{\tt \begin{tabbing} EXPR PROCEDURE MAPCAN(X, FN); \\
\hspace*{1em} IF\= NULL X THEN NIL \\
\> ELSE NCONC(FN CAR X, MAPCAN(CDR X, FN));
\end{tabbing}}}


\de{MAPCAR}{(X:list, FN:function):\ty{any}}{eval, spread}
{Returned is a constructed list of FN applied to each CAR of list X.

{\tt \begin{tabbing} EXPR PROCEDURE MAPCAR(X, FN); \\
\hspace*{1em} IF\= NULL X THEN NIL \\
\> ELSE FN CAR X . MAPCAR(CDR X, FN);
\end{tabbing}}}


\de{MAPCON}{(X:list, FN:function):\ty{any}}{eval, spread}
{Returned is a concatenated list of FN applied to successive CDR
segments of X.

{\tt \begin{tabbing} EXPR PROCEDURE MAPCON(X, FN); \\
\hspace*{1em} IF\= NULL X THEN NIL \\
\> ELSE NCONC(FN X, MAPCON(CDR X, FN));
\end{tabbing}}}


\de{MAPLIST}{(X:list, FN:function):\ty{any}}{eval, spread}
{Returns a constructed list of FN applied to successive CDR segments
of X.

{\tt \begin{tabbing} EXPR PROCEDURE MAPLIST(X, FN); \\
\hspace*{1em} IF\= NULL X THEN NIL \\
\> ELSE FN X . MAPLIST(CDR X, FN);
\end{tabbing}}}


\subsection{Composite Functions}

\de{APPEND}{(\p{U}:\ty{list}, \p{V}:\ty{list}):\ty{list}}{eval, spread}
{Returns a constructed list in which the last element of U is followed
by the first element of V. The list U is copied, V is not.

{\tt \begin{tabbing} EXPR PROCEDURE APPEND(U, V); \\
\hspace*{1em} IF\= NULL U THEN V \\
\> ELSE CAR U . APPEND(CDR U, V);
\end{tabbing}}}

\de{ASSOC}{(\p{U}:\ty{any}, \p{V}:\ty{alist}):\{\ty{dotted-pair},
NIL\}}{eval, spread}
{If U occurs as the CAR portion of an element of the alist V, the
dotted-pair in which U occurred is returned, else NIL is returned.
ASSOC might not detect a poorly formed alist so an invalid
\index{EQUAL ! in ASSOC} \index{alist ! in ASSOC}
construction may be detected by CAR or CDR.

{\tt \begin{tabbing} EXPR PROCEDURE ASSOC(U, V); \\
\hspace*{1em} IF \= NULL V THEN NIL \\
\> ELSE \= IF ATOM CAR V THEN \\
\> \> ERROR(000, LIST(V, "is a poorly formed alist")) \\
\> ELSE IF U = CAAR V THEN CAR V \\
\> ELSE ASSOC(U, CDR V);
\end{tabbing}}
}

\de{DEFLIST}{(\p{U}:\ty{dlist}, \p{IND}:\ty{id}):\ty{list}}{eval, spread}
{A "dlist" is a list in which each element is a two element list:
\index{dlist}
(ID:id PROP:any). Each ID in U has the indicator IND with property
PROP placed on its property list by the PUT function. The value of
DEFLIST is a list of the first elements of each two element list.
Like PUT, DEFLIST may not be used to define functions.

{\tt \begin{tabbing} EXPR PROCEDURE DEFLIST(U, IND); \\
\hspace*{1em} IF NULL U THEN NIL \\
\hspace*{2em} ELSE $<<$ \= PUT(CAAR U, IND, CADAR U); \\
\> CAAR U $>>$ . DEFLIST(CDR U, IND);
\end{tabbing}}
}

\de{DELETE}{(\p{U}:\ty{any}, \p{V}:\ty{list}):\ty{list}}{eval, spread}
{Returns V with the first top level occurrence of U removed from it.
\index{EQUAL ! in DELETE}

{\tt \begin{tabbing} EXPR PROCEDURE DELETE(U, V); \\
\hspace*{1em} IF NULL V THEN NIL \\
\hspace*{2em} ELSE IF CAR V = U THEN CDR V \\
\hspace*{2em} ELSE CAR V . DELETE(U, CDR V);
\end{tabbing}}}

\de{DIGIT}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is a digit, otherwise NIL.

{\tt \begin{tabbing} EXPR PROCEDURE DIGIT(U); \\
\hspace*{1em} IF MEMQ(U, '(!0 !1 !2 !3 !4 !5 !6 !7 !8 !9)) \\
\hspace*{2em} THEN T ELSE NIL;
\end{tabbing}}}

\de{LENGTH}{(\p{X}:\ty{any}):\ty{integer}}{eval, spread}
{The top level length of the list X is returned.

{\tt \begin{tabbing} EXPR PROCEDURE LENGTH(X); \\
\hspace*{1em} IF ATOM X THEN 0 \\
\hspace*{2em} ELSE PLUS(1, LENGTH CDR X);
\end{tabbing}}}

\de{LITER}{(\p{U}:\ty{any}):\ty{boolean}}{eval, spread}
{Returns T if U is a character of the alphabet, NIL
otherwise.\footnote{The published report omits escape characters.
These are required for both upper and lower case as some systems
default to lower.}

{\tt \begin{tabbing} EXPR PROCEDURE LITER(U); \\
\hspace*{1em} IF \= MEMQ(U, '(\=!A !B !C !D !E !F !G !H !I !J !K !L !M \\
\> \> !N !O !P !Q !R !S !T !U !V !W !X !Y !Z \\
\> \> !a !b !c !d !e !f !g !h !i !j !k !l !m \\
\> \> !n !o !p !q !r !s !t !u !v !w !x !y !z)) \\
\> THEN T ELSE NIL;
\end{tabbing}}}

\de{MEMBER}{(\p{A}:\ty{any}, \p{B}:\ty{list}):\ty{extra-boolean}}{eval, spread}
{Returns NIL if A is not a member of list B, returns the remainder of
B whose first element is A. \index{EQUAL ! in MEMBER}

{\tt \begin{tabbing} EXPR PROCEDURE MEMBER(A, B); \\
\hspace*{1em} IF NULL B THEN NIL \\
\hspace*{2em} ELSE IF A = CAR B THEN B \\
\hspace*{2em} ELSE MEMBER(A, CDR B);
\end{tabbing}}}


\de{MEMQ}{(\p{A}:\ty{any}, \p{B}:\ty{list}):\ty{extra-boolean}}{eval, spread}
{Same as MEMBER but an EQ check is used for comparison. \index{EQ ! in
MEMQ}

{\tt \begin{tabbing} EXPR PROCEDURE MEMQ(A, B); \\
\hspace*{1em} IF \= NULL B THEN NIL \\
\> ELSE IF A EQ CAR B THEN B \\
\> ELSE MEMQ(A, CDR B);
\end{tabbing}}}

\de{NCONC}{(\p{U}:\ty{list}, \p{V}:\ty{list}):\ty{list}}{eval, spread}
{Concatenates V to U without copying U. The last CDR of U is modified
to point to V.

{\tt \begin{tabbing} EXPR PROCEDURE NCONC(U, V); \\ BEGIN SCALAR W; \\
\hspace*{2em} \= IF NULL U THEN RETURN V; \\
\> W := U; \\
\> WHILE CDR W DO W := CDR W; \\
\> RPLACD(W, V); \\
\> RETURN U \\
END;
\end{tabbing}}}

\de{PAIR}{(\p{U}:\ty{list}, \p{V}:\ty{list}):\ty{alist}}{eval, spread}
{U and V are lists which must have an identical number of elements. If
not, an error occurs (the 000 used in the ERROR call is arbitrary and
need not be adhered to). Returned is a list where each element is a
dotted-pair, the CAR of the pair being from U, and the CDR the
corresponding element from V.

{\tt \begin{tabbing} EXPR PROCEDURE PAIR(U, V); \\
\hspace*{1em} IF AND(U, V) THEN (CAR U . CAR V) . PAIR(CDR U, CDR V) \\
\hspace*{2em} \= ELSE IF OR(U, V) THEN ERROR(000, \\
\hspace*{4em} "Different length lists in PAIR") \\
\> ELSE NIL;
\end{tabbing}}}


\de{REVERSE}{(\p{U}:\ty{list}):\ty{list}}{eval, spread}
{Returns a copy of the top level of U in reverse order.

{\tt \begin{tabbing} EXPR PROCEDURE REVERSE(U); \\ BEGIN SCALAR W; \\
\hspace*{2em} \= WHILE U DO $<<$ \= W := CAR U . W; \\
\> \> U := CDR U $>>$; \\
\>  RETURN W \\
END;
\end{tabbing}}}

\de{SASSOC}{(\p{U}:\ty{any}, \p{V}:\ty{alist},
\p{FN}:\ty{function}):\ty{any}}{eval, spread}
{Searches the alist V for an occurrence of U. If U is not in the alist
the evaluation of function FN is returned. \index{EQUAL ! in SASSOC}
\index{alist ! in SASSOC}

{\tt \begin{tabbing} EXPR PROCEDURE SASSOC(U, V, FN); \\
\hspace*{1em} IF NULL V THEN FN() \\
\hspace*{2em} \= ELSE IF U = CAAR V THEN CAR V \\
\> ELSE SASSOC(U, CDR V, FN);
\end{tabbing}}}

\de{SUBLIS}{(\p{X}:\ty{alist}, \p{Y}:\ty{any}):\ty{any}}{eval, spread}
{The value returned is the result of substituting the CDR of each
element of the alist X for every occurrence of the CAR part of that
element in Y. \index{alist ! in SUBLIS}

{\tt \begin{tabbing} EXPR PROCEDURE SUBLIS(X, Y); \\
\hspace*{1em}IF NULL X THEN Y \\
\hspace*{2em} ELSE BEGIN \= SCALAR U; \\
\> U := ASSOC(Y, X); \\
\> RETURN \= IF U THEN CDR U \\
\> \> ELSE IF ATOM Y THEN Y \\
\> \> ELSE \= SUBLIS(X, CAR Y) . \\
\> \> \> SUBLIS(X, CDR Y) \\
\> END;
\end{tabbing}}}

\de{SUBST}{(\p{U}:\ty{any}, \p{V}:\ty{any}, \p{W}:\ty{any}):\ty{any}}{eval,
spread}
{The value returned is the result of substituting U for all
occurrences of V in W. \index{EQUAL ! in SUBST}

{\tt \begin{tabbing} EXPR PROCEDURE SUBST(U, V, W); \\
\hspace*{1em} IF NULL W THEN NIL \\
\hspace*{2em} \= ELSE IF V = W THEN U \\
\> ELSE IF ATOM W THEN W \\
\> ELSE SUBST(U, V, CAR W) . SUBST(U, V, CDR W);
\end{tabbing}}}


\subsection{The Interpreter}
\label{interpreter}
\de{APPLY}{(\p{FN}:\{\ty{id,function}\},
\p{ARGS}:\ty{any-list}):\ty{any}}{eval, spread}
{APPLY returns the value of FN with actual parameters ARGS. The actual
parameters in ARGS are already in the form required for binding to the
formal parameters of FN. Implementation specific portions described in
English are enclosed in boxes.

{\tt \begin{tabbing} EXPR PROCEDURE APPLY(FN, ARGS); \\ BEGIN SCALAR
DEFN; \\
\hspace*{2em}\= IF CODEP FN THEN RETURN \\
\> \hspace{1em} \framebox[3.25in]{\parbox{3.25in}{Spread the actual
parameters in ARGS
following the conventions: for calling functions, transfer to the
entry point of the function, and return the value returned by the
function.}}; \\
\> IF \= IDP FN THEN RETURN \\
\> \> IF \= NULL(DEFN := GETD FN) THEN \\
\> \> \> ERROR(000, LIST(FN, "is an undefined function")) \\
\> \> ELSE IF CAR DEFN EQ 'EXPR THEN \\
\> \> \> APPLY(CDR DEFN, ARGS) \\
\> \> ELSE ERROR(000, \\
\> \> \> LIST(FN, "cannot be evaluated by APPLY")); \\
\> IF OR(ATOM FN, NOT(CAR FN EQ 'LAMBDA)) THEN \\
\> \> ERROR(000, \\
\> \> LIST(FN, "cannot be evaluated by APPLY")); \\
\> RETURN \\
\> \> \framebox[3.25in]{\parbox{3.25in}{Bind the actual parameters in ARGS to
the formal
parameters of the lambda expression. If the two lists are not of equal
length then ERROR(000, "Number of parameters do not match"); The value
returned is EVAL CADDR FN.}} \\ END;
\end{tabbing}}}

\de{EVAL}{(\p{U}:\ty{any}):\ty{any}}{eval, spread}
{The value of the expression U is computed. Error numbers are
arbitrary. Portions of EVAL involving machine specific coding are
expressed in English enclosed in boxes.

{\tt \begin{tabbing} EXPR PROCEDURE EVAL(U); \\ BEGIN SCALAR FN; \\
\hspace*{2em} \= IF CONSTANTP U THEN RETURN U; \\
\> IF IDP U THEN RETURN \\
\> \hspace{1em} \framebox[3.25in]{\parbox{3.25in}{U is an id. Return the
value most currently
bound to U or if there is no such binding: ERROR(000, LIST("Unbound:",
U));}} \\
\> IF \= PAIRP CAR U THEN RETURN \\
\> \> IF CAAR U EQ 'LAMBDA THEN APPLY(CAR U, EVLIS CDR U) \\
\> \> ELSE ERROR(\= 000, LIST(CAR U, \\
\> \> \> "improperly formed LAMBDA expression")) \\
\> \> ELSE IF CODEP CAR U THEN \\
\> \> \> RETURN APPLY(CAR U, EVLIS CDR U); \\
\> FN := GETD CAR U; \\
\> IF NULL FN THEN \\
\> \> ERROR(000, LIST(CAR U, "is an undefined function")) \\
\> ELSE IF CAR FN EQ 'EXPR THEN \\
\> \> RETURN APPLY(CDR FN, EVLIS CDR U) \\
\> ELSE IF CAR FN EQ 'FEXPR THEN \\
\> \> RETURN APPLY(CDR FN, LIST CDR U) \\
\> ELSE IF CAR FN EQ 'MACRO THEN \\
\> \> RETURN EVAL APPLY(CDR FN, LIST U) \\
END;
\end{tabbing}}}

\de{EVLIS}{(\p{U}:\ty{any-list}):\ty{any-list}}{eval, spread}
{EVLIS returns a list of the evaluation of each element of U.

{\tt \begin{tabbing} EXPR PROCEDURE EVLIS(U); \\
\hspace*{1em} IF NULL U THEN NIL \\
\hspace*{2em} ELSE EVAL CAR U . EVLIS CDR U;
\end{tabbing}}}

\de{EXPAND}{(\p{L}:\ty{list}, \p{FN}:\ty{function}):\ty{list}}{eval, spread}
{FN is a defined function of two arguments to be used in the expansion
of a MACRO. EXPAND returns a list in the form:

\vspace{.15in}
(FN L$_0$ (FN L$_1$ \ldots (FN L$_{n-1}$ L$_n$) \ldots ))
\vspace{.15in}

where $n$ is the number of elements in L, L$_i$ is the $i$th element
of L.

{\tt \begin{tabbing} EXPR PROCEDURE EXPAND(L,FN); \\
\hspace*{1em} IF NULL CDR L THEN CAR L \\
\hspace*{2em} ELSE LIST(FN, CAR L, EXPAND(CDR L, FN));
\end{tabbing}}}

\de{FUNCTION}{(\p{FN}:\ty{function}):\ty{function}}{noeval, nospread}
{The function FN is to be passed to another function. If FN is to have
side effects its free variables must be fluid or global. FUNCTION is
like QUOTE but its argument may be affected by compilation. We do not
\index{FUNARGs not supported}
consider FUNARGs in this report.}


\de{QUOTE}{(U:any):\ty{any}}{noeval, nospread}
{Stops evaluation and returns U unevaluated.

{\tt \begin{tabbing} FEXPR PROCEDURE QUOTE(U); \\
\hspace*{2em}CAR U;
\end{tabbing}}}

\subsection{Input and Output}
\label{IO}
The user normally communicates with Standard LISP through
\index{standard devices}
``standard devices''. The default devices are selected in accordance
with the conventions of the implementation site. Other input and
output devices or files may be selected for reading and writing using
the functions described herein.



\de{CLOSE}{(\p{FILEHANDLE}:\ty{any}):\ty{any}}{eval, spread}
{Closes the file with the internal name FILEHANDLE writing any
necessary end of file marks and such. The value of FILEHANDLE is that
returned by the corresponding OPEN. \index{OPEN} The value returned is
the value of FILEHANDLE. An error occurs if the file can not be
\index{file handle} \index{files}
closed.

\errormessage{   ***** FILEHANDLE could not be closed}
}

\de{EJECT}{():NIL}{eval, spread}
{Skip to the top of the next output page. Automatic EJECTs are
executed by the print functions when the length set by the PAGELENGTH
\index{PAGELENGTH} function is exceeded.}


\de{LINELENGTH}{(\p{LEN}:\{\ty{integer}, NIL\}):\ty{integer}}{eval, spread}
{If LEN is an integer the maximum line length to be printed before the
print functions initiate an automatic TERPRI is set to the value LEN.
\index{TERPRI}
No initial Standard LISP line length is assumed. The previous line
length is returned except when LEN is NIL. This special case returns
the current line length and does not cause it to be reset. An error
occurs if the requested line length is too large for the currently
selected output file or LEN is negative or zero.

\errormessage{   ***** LEN is an invalid line length}
}


\de{LPOSN}{():\ty{integer}}{eval, spread}
{Returns the number of lines printed on the current page. At the top
of a page, 0 is returned. }


\de{OPEN}{(\p{FILE}:\ty{any}, \p{HOW}:\ty{id}):\ty{any}}{eval, spread}
{Open the file with the system dependent name FILE for output if HOW
is EQ to OUTPUT, or input if HOW is EQ to INPUT. If the file is
\index{file handle} \index{files} \index{OUTPUT} \index{INPUT}
opened successfully, a value which is internally associated with the
file is returned. This value must be saved for use by RDS and WRS. An
error occurs if HOW is something other than INPUT or OUTPUT or the
file can't be opened.

\errormessage{***** HOW is not option for OPEN}
\errormessage{***** FILE could not be opened}
}


\de{PAGELENGTH}{(\p{LEN}:\{\ty{integer}, NIL\}):\ty{integer}}{eval, spread}
{Sets the vertical length (in lines) of an output page. Automatic page
EJECTs are executed by the print functions when this length is
\index{EJECT}
reached. The initial vertical length is implementation specific. The
previous page length is returned. If LEN is 0, no automatic page
ejects will occur. }


\de{POSN}{():\ty{integer}}{eval, spread}
{Returns the number of characters in the output buffer. When the
buffer is empty, 0 is returned.}


\de{PRINC}{(\p{U}:\ty{id}):\ty{id}}{eval, spread}
{U must be a single character id such as produced by EXPLODE or read
by READCH or the value of !\$EOL!\$. The effect is the character U
\index{\$EOL\$ (global)}
displayed upon the currently selected output device. The value of
!\$EOL!\$ causes termination of the current line like a call to
TERPRI.}


\de{PRINT}{(\p{U}:\ty{any}):\ty{any}}{eval, spread}
{Displays U in READ readable format and terminates the print line. The
value of U is returned.

{\tt \begin{tabbing} EXPR PROCEDURE PRINT(U); \\
\hspace*{2em} $<<$ PRIN1 U; TERPRI(); U $>>$;
\end{tabbing}}}


\de{PRIN1}{(\p{U}:\ty{any}):\ty{any}}{eval, spread}
{U is displayed in a READ readable form. The format of display is the
result of EXPLODE expansion; special characters are prefixed with the
escape character !, and strings are enclosed in "\ldots ". Lists are
displayed in list-notation and vectors in vector-notation. }


\de{PRIN2}{(\p{U}:\ty{any}):\ty{any}}{eval, spread}
{U is displayed upon the currently selected print device but output is
not READ readable. The value of U is returned. Items are displayed as
described in the EXPLODE function with the exceptions that the escape
character does not prefix special characters and strings are not
enclosed in "\ldots ". Lists are displayed in list-notation and
vectors in vector-notation. The value of U is returned. }


\de{RDS}{(\p{FILEHANDLE}:\ty{any}):\ty{any}}{eval, spread}
{Input from the currently selected input file is suspended and further
input comes from the file named. FILEHANDLE is a system dependent
\index{file handle}
internal name which is a value returned by OPEN. If FILEHANDLE is NIL
the standard input device is selected. When end of file is reached on
a non-standard input device, the standard input device is reselected.
When end of file occurs on the standard input device the Standard LISP
reader terminates. RDS returns the internal name of the previously
selected input file.
\index{standard input}

\errormessage{***** FILEHANDLE could not be selected for input}
}


\de{READ}{():\ty{any}}{}
{The next expression from the file currently selected for input. Valid
input forms are: vector-notation, dot-notation, list-notation,
numbers, function-pointers, strings, and identifiers with escape
characters. Identifiers are interned onW the OBLIST (see
\index{INTERN} \index{OBLIST entry}
the INTERN function in "Identifiers", section~\ref{identifiers} on
page~\pageref{identifiers}). READ returns the
\index{\$EOF\$ (global)}
value of !\$EOF!\$ when the end of the currently selected input file
is reached. }


\de{READCH}{():\ty{id}}{}
{Returns the next interned character from the file currently selected
for input. Two special cases occur. If all the characters in an input
\index{\$EOL\$ (global)} \index{\$EOF\$ (global)} record have been read,
the value of !\$EOL!\$ is returned. If the file selected for input has
all been read the value of !\$EOF!\$ is returned. Comments delimited
by \% and end-of-line are not transparent to READCH. \index{\% ! read
by READCH} }


\de{TERPRI}{():\p{NIL}}{}
{The current print line is terminated.}


\de{WRS}{(\p{FILEHANDLE}:\ty{any}):\ty{any}}{eval, spread}
{Output to the currently active output file is suspended and further
output is directed to the file named. FILEHANDLE is an internal name
which is returned by OPEN. The file named must have been opened for
output. If FILEHANDLE is NIL the standard output device is selected.
\index{file handle} \index{standard output}
WRS returns the internal name of the previously selected output file.

\errormessage{***** FILEHANDLE could not be selected for output}
}

\subsection{LISP Reader}

An EVAL read loop has been chosen to drive a Standard LISP system to
provide a continuity in functional syntax. Choices of messages and the
amount of extra information displayed are decisions left to the
implementor.

\index{STANDARD-LISP}
{\tt \begin{tabbing} EXPR PROCEDURE STANDARD!-LISP(); \\ BEGIN SCALAR
VALUE; \\
\hspace*{2em} \= RDS NIL;  WRS NIL; \\
\> PRIN2 "Standard LISP"; TERPRI(); \\
\> WHILE T DO \\
\> \hspace*{1em} $<<$ \= PRIN2 "EVAL:"; TERPRI(); \\
\> \> VALUE := ERRORSET(QUOTE EVAL READ(), T, T); \\
\> \> IF NOT ATOM VALUE THEN PRINT CAR VALUE; \\
\> \> TERPRI() $>>$; \\
END;
\end{tabbing}}

\de{QUIT}{()}{}
{Causes termination of the LISP reader and control to be transferred
to the operating system.}

\section{System GLOBAL Variables}
\label{slglobals}

These variables provide global control of the LISP system, or
implement values which are constant throughout execution.\footnote{The
published document does not specify that all these are GLOBAL.}


\variable{*COMP}{NIL}{global}
{The value of !*COMP controls whether or not PUTD compiles the
function defined in its arguments before defining it. If !*COMP is NIL
the function is defined as an xEXPR. If !*COMP is something else the
function is first compiled. Compilation will produce certain changes
in the semantics of functions particularly FLUID type access.}


\variable{EMSG*}{NIL}{global}
{Will contain the MESSAGE generated by the last ERROR call (see
\index{ERROR}
``Error Handling'' section~\ref{errors} on page~\pageref{errors}).}


\variable{\$EOF\$}{\s{an uninterned identifier}}{global}
{The value of !\$EOF!\$ is returned by all input functions when the
end
\index{end of file}
of the currently selected input file is reached.}


\variable{\$EOL\$}{\s{an uninterned identifier}}{global}
{The value of !\$EOL!\$ is returned by READCH when it reaches the end
of
\index{READCH} \index{end of line} \index{PRINC}
a logical input record. Likewise PRINC will terminate its current line
(like a call to TERPRI) when !\$EOL!\$ is its argument.}

\variable{*GC}{NIL}{global}
{!*GC controls the printing of garbage collector messages.  If NIL no
\index{garbage collector}
indication of garbage collection may occur.  If non-NIL various system
dependent messages may be displayed.}


\variable{NIL}{NIL}{global}
{NIL is a special global variable. It is protected from being modified
by SET or SETQ.
\index{NIL ! cannot be changed}}


\variable{*RAISE}{NIL}{global}
{If !*RAISE is non-NIL all characters input through Standard LISP
input/output functions will be raised to upper case. If !*RAISE is NIL
characters will be input as is.}


\variable{T}{T}{global}
{T is a special global variable. It is protected from being modified
by SET or SETQ. \index{T ! cannot be changed}}


\section{The Extended Syntax}

Whenever it is possible to define Standard LISP functions in LISP the
text of the function will appear in an extended syntax.  These
definitions are supplied as an aid to understanding the behavior of
functions and not as a strict implementation guide.  A formal scheme
for the translation of extended syntax to Standard LISP is presented
to eliminate misinterpretation of the definitions.

\subsection{Definition}
The goal of the transformation scheme is to produce a PUTD invocation
which has the function translated from the extended syntax as its
actual parameter.  A rule has a name in brackets
\s{\ldots} by which it is known and is defined by what follows the meta 
symbol ::=.  Each rule of the set consists of one or more
``alternatives'' separated by the $\mid$ meta symbol, being the
different ways in which the rule will be matched by source text.  Each
alternative is composed of a ``recognizer'' and a ``generator''
separated by the $\Longrightarrow$ meta symbol.  The recognizer is a
concatenation of any of three different forms.  1) Terminals - Upper
case lexemes and punctuation which is not part of the meta syntax
represent items which must appear as is in the source text for the
rule to succeed.  2) Rules - Lower case lexemes enclosed in \s{\ldots}
are names of other rules.  The source text is matched if the named
rule succeeds.  3) Primitives - Lower case singletons not in brackets
are names of primitives or primitive classes of Standard LISP.  The
syntax and semantics of the primitives are given in Part I.

The recognizer portion of the following rule matches an extended
syntax procedure:
 

\s{function} ::= ftype PROCEDURE id (\s{id list});  \\
\hspace*{2em} \s{statement}; $\Longrightarrow$
 
A function is recognized as an ``ftype'' (one of the tokens EXPR,
FEXPR, etc.) followed by the keyword PROCEDURE, followed by an ``id''
(the name of the function), followed by an \s{id list} (the formal
parameter names) enclosed in parentheses.  A semicolon terminates the
title line.  The body of the function is a
\s{statement} followed by a semicolon.  For example: 
 
\begin{verbatim}
EXPR PROCEDURE NULL(X); EQ(X, NIL);
\end{verbatim}

\noindent satisfies the recognizer, causes the generator to be activated and 
the rule to be matched successfully.

The generator is a template into which generated items are
substituted.  The three syntactic entities have corresponding meanings
when they appear in the generator portion.  1) Terminals - These
lexemes are copied as is to the generated text.  2) Rules - If a rule
has succeeded in the recognizer section then the value of the rule is
the result of the generator portion of that rule.  3) Primitives -
When primitives are matched the primitive lexeme replaces its
occurrence in the generator.
 
If more than one occurrence of an item would cause ambiguity in the
generator portion this entity appears with a bracketed subscript.
Thus:
 
\begin{tabbing}
\s{conditional} ::= \\
\hspace*{2em} IF \s{expression} \= THEN \s{statement$_1$} \\ 
\> ELSE \s{statement$_2$} \ldots
\end{tabbing}
 
\noindent has occurrences of two different \s{statement}s.  The generator 
portion uses the subscripted entities to reference the proper
generated value.

The \s{function} rule appears in its entirety as:

\begin{tabbing}
\s{function} ::= ftype PROCEDURE id (\s{id list});\s{statement};
$\Longrightarrow$ \\
\hspace*{2em} \=(PUTD \= (QUOTE id) \\
\> \> (QUOTE ftype) \\
\> \>(QUOTE (LAMBDA (\s{id list}) \s{statement})))
\end{tabbing}
 
If the recognizer succeeds (as it would in the case of the NULL
procedure example) the generator returns:

\begin{verbatim}
(PUTD (QUOTE NULL) (QUOTE EXPR) (QUOTE (LAMBDA (X) (EQ X NIL))))
\end{verbatim}
 
The identifier in the template is replaced by the procedure name NULL,
\s{id list} by the single formal parameter X, the \s{statement} by (EQ
X NIL) which is the result of the \s{statement} generator.  EXPR
replaces ftype, the type of the defined procedure.
 
 
\subsection{The Extended Syntax Rules}

\begin{tabbing}
\s{function} ::= ftype \k{PROCEDURE} id (\s{id list}); \s{statement};
$\Longrightarrow$ \\
\hspace*{2em} \= (PUTD \= (QUOTE id) \\
\> \> (QUOTE ftype) \\
\> \> (QUOTE (LAMBDA (\s{id list}) \s{statement}))) \\ \\

\s{id list} ::= id $\Longrightarrow$ id $\mid$ \\
\> id, \s{id list} $\Longrightarrow$ id \s{id list} $\mid$ \\
\> $\Longrightarrow$ NIL \\

\s{statement} ::= \s{expression} $\Longrightarrow$ \s{expression} $\mid$ \\
\> \s{proper statement} $\Longrightarrow$ \s{proper statement} \\ \\

\s{proper statement} ::=  \\
\> \s{assignment statement} $\Longrightarrow$ \s{assignment statement}
$\mid$ \\
\> \s{conditional statement} $\Longrightarrow$ \s{conditional statement}
$\mid$ \\
\> \s{while statement} $\Longrightarrow$ \s{while statement} $\mid$ \\
\> \s{compound statement} $\Longrightarrow$ \s{compound statement} \\ \\

\s{assignment statement} ::= id := \s{expression} $\Longrightarrow$ \\
\> \> (SETQ id \s{expression}) \\ \\

\s{conditional statement} ::= \\
\> \k{IF} \s{expression} \k{THEN} \s{statement$_1$} \k{ELSE}
\s{statement$_2$} $\Longrightarrow$ \\
\> \hspace{2em} \= (COND (\s{expression} \s{statement$_1$})(T
\s{statement$_2$})) $\mid$ \\
\> \k{IF} \s{expression} \k{THEN} \s{statement} $\Longrightarrow$ \\
\> \> (COND (\s{expression} \s{statement})) \\ \\

\s{while statement} ::= \k{WHILE} \s{expression} \k{DO} \s{statement}
$\Longrightarrow$ \\
\> \> (PROG NIL \\
\> \> LBL \= (COND ((NULL \s{expression}) (RETURN NIL))) \\
\> \> \> \s{statement} \\
\> \> \> (GO LBL))  \\ \\

\s{compound statement} ::= \\
\> \k{BEGIN} \k{SCALAR} \s{id list}; \s{program list} \k{END}
$\Longrightarrow$ \\
\> \> (PROG (\s{id list}) \s{program list}) $\mid$ \\
\> \k{BEGIN} \s{program list} \k{END} $\Longrightarrow$ \\
\> \> (PROG NIL \s{program list}) $\mid$ \\
\> \k{$<<$} \s{statement list} \k{$>>$} $\Longrightarrow$ (PROGN
\s{statement list}) \\ \\

\s{program list} ::= \s{full statement} $\Longrightarrow$ \s{full statement}
 $\mid$ \\
\> \s{full statement} \s{program list} $\Longrightarrow$ \\
\> \> \s{full statement} \s{program list} \\ \\

\s{full statement} ::= \s{statement} $\Longrightarrow$ \s{statement} $\mid$
id: $\Longrightarrow$ id  \\ \\

\s{statement list} ::= \s{statement} $\Longrightarrow$ \s{statement} $\mid$ \\
\> \s{statement}; \s{statement list} $\Longrightarrow$ \\
\> \> \s{statement} \s{statement list}  \\ \\

\s{expression} ::= \\
\> \s{expression$_1$} \k{.} \s{expression$_2$} $\Longrightarrow$ \\
\> \> (CONS \s{expression$_1$} \s{expression$_2$} $\mid$ \\
\> \s{expression$_1$} \k{=}  \s{expression$_2$} $\Longrightarrow$ \\
\> \> (EQUAL \s{expression$_1$} \s{expression$_2$}) $\mid$ \\
\> \s{expression$_1$} \k{EQ} \s{expression$_2$} $\Longrightarrow$ \\
\> \> (EQ \s{expression$_1$} \s{expression$_2$}) $\mid$ \\
\> '\s{expression} $\Longrightarrow$ (QUOTE \s{expression}) $\mid$ \\
\> function \s{expression} $\Longrightarrow$ (function \s{expression})
$\mid$ \\
\> function(\s{argument list}) $\Longrightarrow$ (function \s{argument list})
$\mid$ \\
\>  number $\Longrightarrow$ number $\mid$ \\
\> id $\Longrightarrow$ id \\ \\

\s{argument list} ::= () $\Longrightarrow$ $\mid$ \\
\> \s{expression} $\Longrightarrow$ \s{expression} $\mid$ \\
\> \s{expression}, \s{argument list} $\Longrightarrow$ \s{expression}
\s{argument list}
\end{tabbing}
 
Notice the three infix operators .  EQ and = which are translated into
calls on CONS, EQ, and EQUAL respectively.  Note also that a call on a
function which has no formal parameters must have () as an argument
list.  The QUOTE function is abbreviated by '.
\bibliography{sl}
\bibliographystyle{plain}
\end{document}
